The EDHMM captures the relationships among state $x_t$, duration $d_t$, and  observation $y_t$ over time $t$. It consists of four components: the initial state distribution, the transition distributions, the observation distributions, and the duration distributions. 

We define the observation sequence $\mathcal{Y} = \{y_1, y_2, \ldots, y_T\}$; the latent state sequence $\mathcal{X} = \{x_0, x_1, x_2, \ldots, x_T\}$; and the remaining time in each segment $\mathcal{D} = \{ d_1, d_2, \ldots, d_T\}$, where $x_t \in \{ 1, 2, \ldots, K\}$ with $K$ the maximum number of states, $d_t \in \{1, 2, \ldots \}$, and $y_t \in \mathbb{R}^n$.     We assume that the Markov chain on the latent states is homogenous, i.e., that $p(x_t = i | x_{t-1}=j, A) = a_{i,j} \forall t$ where $A$ is a $K\times K$ matrix with element $a_{i,j}$ at row $i$ and column $j.$  The prior on $A$ is row-wise Dirichlet with zero prior mass on self-transitions, i.e.  $p(a_{i,:}) = \mathrm{Dir}({1}/{K-1}, \ldots, 0 , \ldots {1}/{K-1})$ where $a_{i,:}$ is a row vector and the $i$th Dirichlet parameter is $0.$  Each state is imbued with its own duration distribution $p(d_t | x_t = k) = p(d_t | \lambda_{k})$ with parameter $\lambda_k$.  Each duration distribution parameter is drawn from a prior $p(\lambda_k)$ which can be chosen in an application specific way.  The collection of all duration distribution parameters is $\lambda = \{\lambda_1, \ldots, \lambda_K\}$.  Each state is also imbued with an observation generating distribution $p(y_t  | x_t = k) = p(y_t | \theta_{k})$ with parameter $\theta_k$.  Each observation distribution parameter is 
%given 
drawn from a prior $p(\theta_k)$ also to be chosen according to the application.   The set of all observation distribution parameters is $\theta.$  In the following exposition, explicit conditional dependencies on component distribution parameters are omitted to focus on the particulars unique to the EDHMM.

In an EDHMM the transitions between states are only allowed at the end of a segment:
    \begin{equation}
        p(x_t | x_{t-1}, d_{t-1}) = 
        \begin{cases} 
            \delta(x_t, x_{t-1}) & \textrm{if $d_{t-1} > 1$} \\
            p(x_t | x_{t-1}) & \textrm{otherwise}
        \end{cases}
    \end{equation}
where the Kronecker delta $\delta(a,b) = 1$ if $a=b$ and zero otherwise. The duration distribution generates segment lengths at every state switch:
    \begin{equation}
        p(d_t | x_{t}, d_{t-1}) = 
        \begin{cases} 
            \delta(d_t, d_{t-1}-1) & \textrm{if $d_{t-1} > 1$} \\
            p(d_t | x_{t}) & \textrm{otherwise.}
        \end{cases}
    \end{equation}
The joint distribution of the EDHMM is 
\begin{multline}
    \label{eq:joint}
    p(\mathcal{X},\mathcal{D},\mathcal{Y}) =  \\ 
    p(x_0)p(d_0)\prod_{t=1}^T p(y_t | x_t, \theta) p(x_t | x_{t-1}, d_{t-1}, A) p(d_t | x_{t}, d_{t-1}, \lambda)
\end{multline}
corresponding to the graphical model in Figure \ref{fig:graphical model}. Alternative choices to define the duration variable $d_t$ exist; see \cite{Chiappa2011} for details. Algorithm \ref{alg:gen} illustrates the EDHMM as a generative model.